Exponential methods for solving hyperbolic problems with application to kinetic equations

### 2020-12-01 communication/20-12-04_canum/canum.tex
🖥  final version of canum slides ?

@@

  Linear part is exact! \cmark
\end{frame}
%-------%
\begin{frame}{Idea of exponential integrators}
  \textbf{\color{mblue} 2 classes of methods:}
  \begin{description}
    \item[\textbf{exponential Runge-Kutta:}] solve exactly what we can, and interpolate the rest. For example first order exponential Euler method:
      $$
        u(t_n+\Delta t) \approx u^{n+1} = e^{-ia\Delta t}u^n + \Delta t\varphi_1(ia\Delta t)F(u^n)
      $$
      where $\varphi_1(z) = \dfrac{e^z - 1}{z}$
      \vspace{-0.1cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Hochbruck:2010}
      \end{thebibliography}
    \item[\textbf{Lawson:}] Change of variable: $v(t)=e^{-iat}u(t)$, we solve with a RK method: $\dot{v} = \tilde{F}(t,v) = e^{-iat}F(e^{iat}v(t))$

      For example, Lawson Euler method:
      $$
        v(t_n+\Delta t)\approx v^{n+1} = v^n + \Delta t e^{-iat_n}F(e^{iat_n}v^n)
      $$
      or as an expression of $u$:
      $$
        u^{n+1} = e^{-ia\Delta t}u^n + \Delta te^{ia\Delta t}F(u^n)
      $$
      \vspace{-0.75cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Isherwood:2018}
        %\bibitem{a} \cite{Isherwood:2018}
      \end{thebibliography}
  \end{description}
\end{frame}

@@

  $\lambda$ is the Fourier symbol (or eigenvalues) of FD method to approximate $\partial_vf$.
\end{frame}
%-------%
\begin{frame}{Phase discretization}
  In $v$ direction we use a FD method:
  \begin{itemize}
    \item CD2 (centered difference of order 2): $(\partial_v f)(v_j)\approx \dfrac{f_{j+1}-f_{j-1}}{2\Delta v}$
    \item WENO5 (weighted essentially non-oscillatory of order 5):
      \begin{itemize}
        \item WENO5: non linear scheme: \st{Von Neumann analysis}
        \item LW5 (linearized WENO5): linear scheme (this is Lagrange interpolation of order 5)
      \end{itemize}
      $$
        (\partial_vf)(v_j)\approx\frac{1}{\Delta v}\left(-\frac{1}{30}f_{j-3} + \frac{1}{4}f_{j-2} - f_{j-1} + \frac{1}{3}f_j + \frac{1}{2}f_{j+1} - \frac{1}{20}f_{j+2}\right)
      $$

@@

\subsection{Lawson methods}
%------------------------------------------------------------------------------
\begin{frame}{Lawson methods stability domain}
  For our toy model:
  $$
    \dot{u} = iau + \lambda(u)
  $$
  Change of variable: $v(t) = e^{-iat}u(t)$
  $$
    \dot{v} = e^{-iat}\lambda e^{iat}v
  $$
  Apply a Runge-Kutta method to compute stability function of Lawson method:
  $$
    v^{n+1} = \underbrace{p(\lambda\Delta t)}_{\text{stability function of RK}}v^n
  $$
  \emph{i.e.}:
  $$
    u^{n+1} = \overbrace{p(\lambda\Delta t)e^{-ia\Delta t}}^{\text{stability function of Lawson}}u^n
  $$
  Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$ of Lawson method is \textbf{the same} as the underlying Runge-Kutta method \textbf{because} $ia\in i\mathbb{R}$
\end{frame}
%-------%
\begin{frame}{Considered $Lawson(RK(s,p))$ methods}
  \begin{figure}\centering
    \includegraphics[height=0.8\textheight]{img/rk_sd.png}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2}
  For stability between a Lawson method and CD2, we solve:
  $$
    |p(iy)| = 1,\quad y\in\mathbb{R}
  $$
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_scheme}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_rk44_weno.png}
    \end{column}
  \end{columns}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2/LW5: CFL estimates}
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      - & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      CD2 ($y_{\max}$) & $\sqrt{3}$ & $2\sqrt{2}$\\
      \hline
      LW5 ($\sigma$) & $1.433$   & $1.73$   \\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes.}
  \end{table}
  Other RK--CD2 CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Baldauf:2008}
\end{thebibliography}
  Same estimation of RK--WENO CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Motamed:2010}
  \bibitem{} \customcite{Lunet:2017}
\end{thebibliography}
\end{frame}

@@

  Stability domain depends of $a\Delta t$\dots \xmark
\end{frame}
%-------%
\begin{frame}{}
  \only<1>{
    \begin{figure}\centering
      \includegraphics[height=0.7\textheight]{img/expRK22_sd.png}
      \caption{Stability domain of ExpRK(2,2) for $a\Delta t\in\{1.1, 3.4\}$}
    \end{figure}
  }
  \only<2>{
    \begin{figure}\centering
      \includegraphics[height=0.7\textheight]{img/CM_sd.png}
      \caption{Stability domain of Cox-Matthews for $a\Delta t\in\{1.1, 3.4\}$}
    \end{figure}
  }
\end{frame}

@@

  \mbold{Numerical tools:}
  \begin{itemize}
    \item FFT in $x$ direction
    \item WENO5 in $v$ direction
    \item $Lawson(RK(s,p))$ method in time $t$
  \end{itemize}

>>>>

\usepackage{enumitem}
%\setlist[description]{leftmargin=0.5cm}

@@

  Linear part is exact! \cmark
\end{frame}
%-------%
\begin{frame}{Idea of exponential integrators}
  \textbf{\color{mblue} 2 classes of methods:}
  \begin{description}[leftmargin=0.25cm]
    \item[\mbold{exponential Runge-Kutta:}] solve exactly what we can, and interpolate the rest. For example first order exponential Euler method:
      $$
        u(t_n+\Delta t) \approx u^{n+1} = e^{-ia\Delta t}u^n + \Delta t\varphi_1(ia\Delta t)F(u^n)
      $$
      where $\varphi_1(z) = \dfrac{e^z - 1}{z}$
      \vspace{-0.1cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Hochbruck:2010}
      \end{thebibliography}
    \item[\mbold{Lawson:}] Change of variable: $v(t)=e^{-iat}u(t)$, we solve with a RK method: $\dot{v} = \tilde{F}(t,v) = e^{-iat}F(e^{iat}v(t))$

      For example, Lawson Euler method:
      $$
        v(t_n+\Delta t)\approx v^{n+1} = v^n + \Delta t e^{-iat_n}F(e^{iat_n}v^n)
      $$
      or as an expression of $u$:
      $$
        u^{n+1} = e^{-ia\Delta t}u^n + \Delta te^{ia\Delta t}F(u^n)
      $$
      \vspace{-0.75cm}
      \begin{thebibliography}{9}
        \setbeamertemplate{bibliography item}[article]
        \bibitem{a} \customcite{Isherwood:2018}
      \end{thebibliography}
  \end{description}
\end{frame}

@@

  $\lambda$ is the Fourier symbol (or eigenvalues) of FD method to approximate $\partial_vf$.
\end{frame}
%-------%
\begin{frame}{Phase space discretization}
  In $v$ direction we use a FD method:
  \begin{itemize}
    \item CD2 (centered difference of order 2): $(\partial_v f)(v_j)\approx \dfrac{f_{j+1}-f_{j-1}}{2\Delta v}$
    \item WENO5 (weighted essentially non-oscillatory of order 5):
      \begin{itemize}
        \item WENO5: non linear scheme: \st{Von Neumann analysis}
        \item LW5 (linearized WENO5): linear scheme (this is Lagrange interpolation of order 5)
      \end{itemize}
      $$
        (\partial_vf)(v_j)\approx\frac{1}{\Delta v}\left(-\frac{1}{30}f_{j-3} + \frac{1}{4}f_{j-2} - f_{j-1} + \frac{1}{3}f_j + \frac{1}{2}f_{j+1} - \frac{1}{20}f_{j+2}\right)
      $$

@@

\subsection{Lawson methods}
%------------------------------------------------------------------------------
\begin{frame}{Lawson methods stability domain}
  For our toy model:
  $$
    \dot{u} = iau + \lambda(u)
  $$
  Change of variable: $v(t) = e^{-iat}u(t)$
  $$
    \dot{v} = e^{-iat}\lambda e^{iat}v
  $$
  Apply a Runge-Kutta method to compute stability function of Lawson method:
  $$
    v^{n+1} = \underbrace{p(\lambda\Delta t)}_{\text{stability function of RK}}v^n
  $$
  \emph{i.e.}:
  $$
    u^{n+1} = \overbrace{p(\lambda\Delta t)e^{-ia\Delta t}}^{\text{stability function of Lawson}}u^n
  $$
  Stability domain: $\mathcal{D}=\left\{z\in\mathbb{C},|p(z)|\leq 1\right\}$ of Lawson method is \textbf{the same} as the underlying Runge-Kutta method \textbf{because} $ia\in i\mathbb{R}$
\end{frame}
%-------%
\begin{frame}{Considered $Lawson(RK(s,p))$ methods}
  \begin{figure}\centering
    \includegraphics[height=0.75\textheight]{img/rk_sd.png}
    \caption{$\{z\in\mathbb{C},|p(z)|=1\}$}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2}
  For stability between a Lawson method and CD2, we solve:
  $$
    |p(iy)| = 1,\quad y\in\mathbb{R}
  $$
  \begin{figure}\centering
    \includegraphics[width=0.8\textwidth]{img/yaxis.png}
  \end{figure}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- LW5}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_scheme}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{img/cfl_rk44_weno.png}
    \end{column}
  \end{columns}
\end{frame}
%-------%
\begin{frame}{Lawson methods -- CD2/LW5: CFL estimates}
  \begin{table}
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline
      - & Lawson($RK(3,3)$) & Lawson($RK(4,4)$) \\
      \hline
      CD2 ($y_{\max}$) & $\sqrt{3}$ & $2\sqrt{2}$\\
      \hline
      LW5 ($\sigma$) & $1.433$   & $1.73$   \\
      \hline  
    \end{tabular}
    \caption{CFL number for some Lawson schemes.}
  \end{table}
  Other RK--CD2 CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Baldauf:2008}
\end{thebibliography}
  Same results for RK--WENO CFL estimates:
\begin{thebibliography}{9}
  \setbeamertemplate{bibliography item}[article]
  \bibitem{} \customcite{Motamed:2010}
  \bibitem{} \customcite{Lunet:2017}
\end{thebibliography}
\end{frame}

@@

  Stability domain depends of $a\Delta t$\dots \xmark
\end{frame}
%-------%
\begin{frame}{}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{figure}\centering
        \includegraphics[width=\textwidth]{img/expRK22_sd.png}
        \caption{Stability domain of ExpRK(2,2) for $a\Delta t\in\{1.1, 3.4\}$}
      \end{figure}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{figure}\centering
        \includegraphics[width=\textwidth]{img/CM_sd.png}
        \caption{Stability domain of Cox-Matthews for $a\Delta t\in\{1.1, 3.4\}$}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

@@

  \mbold{Numerical tools:}
  \begin{itemize}
    \item FFT in $x$ direction
    \item CD2 or WENO5 in $v$ direction
    \item $Lawson(RK(s,p))$ method in time $t$
  \end{itemize}
